---
title: "FA3_CUERDO, Naomi Hannah A."
author: "Cuerdo, Naomi Hannah A."
date: "2025-02-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, include=FALSE}
library(ggplot2)
library(tidyverse)
```


```{r dataset}
data("diamonds")
```

1. Create a histogram on the diamonds dataset, for example with
ggplot() + geom_histogram(aes(x = carat), data = diamonds)

Re-write this using the layer function like we did in class. Hint: if you don’t know what the default values for some of the aspects of the plot, examine p$layers.


```{r layer function}

histo_layer <- ggplot(diamonds, aes(x = carat)) +
  layer(
    geom = "bar",
    stat = "bin",
    position = "stack",
    mapping = aes(y = after_stat(count))
  )

histo_layer
```

2. Remember that a histogram is a plot with stat_bin and geom_bar. Modify your histogram code so that it uses a different geom, for example geom_line or geom_point. This should be simple once you have the layer specification of a histogram.

```{r histogram modify using geom_point}

histo_layer2 <- ggplot(diamonds, aes(x = carat)) +
  layer(
    geom = "point",
    stat = "bin",
    position = "identity",
    mapping = aes(y = after_stat(count))
  )

histo_layer2
```

The graph uses points to visualize the data instead of bars. 

We can also use lines for visualization:

```{r histogram modify using geom_line}

histo_layer3 <- ggplot(diamonds, aes(x = carat)) +
  layer(
    geom = "line",
    stat = "bin",
    position = "identity",
    mapping = aes(y = after_stat(count))
  )

histo_layer3

```
3. In your histogram (the one plotted with bars that you created in question 1), add an aesthetic mapping from one of the factor variables (maybe color or clarity) to the fill or color aesthetic.

```{r aesthetic mapping}

histo_layer4 <- ggplot(diamonds, aes(x = carat, fill = color)) +
  layer(
    geom = "bar",
    stat = "bin",
    position = "stack",
    mapping = aes(y = after_stat(count))
  )

histo_layer4
```

4. What is the default position adjustment for a histogram? Try changing the position adjustment in the histogram you created in question 3 to something different (hint: try dodge). 


```{r histo layer 5}

histo_layer5 <- ggplot(diamonds, aes(x = carat, fill = color)) +
  layer(
    geom = "bar",
    stat = "bin",
    position = "dodge",
    mapping = aes(y = after_stat(count))
  )

histo_layer5

```

By default, histograms use **position = "stack"**, but changing it to **dodge** will place the bars next to each other instead of stacking them.

